New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2022-09-12 paper/paper.Rmd
Some work on the software section

XXX copied from the slicing part to be weaved in below:

The current implementation allows the user to individually enter the components of the vector defining the center point, and updates the display by jumping to the new slice.

The current implementation contains a switch to change between the slice view and the projected view of the data, while keeping all other settings fixed.

In practice we define the center point $\mathbf{c}$ explicitly in terms of all $p$ components and calculate the slice display following Eq. \ref{eq:slice}.

This was implemented as a slider, and to allow the user to efficiently explore the interesting range we provide the slider range as an input.

@@

- **Why is this a good sandbox**
- **Explain the functionality available in the notebooks**

Mathematica provides much utility and versatility, such as inbuilt data visualisation, data manipulation and analysis, dynamic functionality, and symbolic and numeric computation. Most of this inbuilt functionality is quite user-friendly and can be found in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there. Also, it is relatively simple  to create dynamic objects via the inbuilt commands `Manipulate` or `DynamicModule`. Control objects, such as sliders, locator panes, and input fields, can then be used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that dynamic variable will be updated. This is the essential ingredient for this implementation of the manual tour. However, it can be hard to alter the behaviour of the control objects such as setting the 1D slider on a different scale  (Slider increments add 10 points into the slice rather than increasing the slice height by 0.1 for example). \textcolor{red}{but this is no longer true if you replace it with animate??}
\textcolor{red}{Alex, is this still true with the controls you are using now?}

The examples presented in the penguins’ notebooks all utilise our primary new function, `SliceDynamic`. This function typically accepts grouped data in the form of a matrix where the second last column details the name of the group and the last column details the group index \textcolor{red}{(which can be one if there is only one group)}. After specifying the initial slice thickness and the slice range, the user is then presented with an interactive display in which the control objects appear on the left and the slice visulisation appears on the right. The user can change the orientation of the slice via the locator pane, which changes the projection matrix;  a slider controls the slice thickness; and there is an input field, which changes the slice centre. The user can also change the appearance of the plot by zooming into the center or changing the point size with sliders provided.\textcolor{red}{this zoom doesn't work very well} The current projection can be displayed to contrast it with the slice by ticking a box. The explicit projection matrix can also be displayed via another checkbox in order to store it for further study.

>>>>

## Structure

The implementation of the manual tour with slice control needs a visualization of the current projection in terms of an axis display. This display should be able to track the mouse position and adjust the projection based on the user selection. The user can select one axis (corresponding to one variable) via clicking on it, and then adjust the position of that axis by dragging it during the click event. In practice the closest axis will be selected, and the dragging results in a tracking of the mouse position that will update the projection in small steps, rendering interpolation unnecessary.

The interaction in the axis display need to be mapped back onto the projection matrix, which in turn needs to be linked to both the orthonormalization procedure and to a second display that is showing the projected data.

In addition, we also want to be able to look at slices of the data and select slicing parameters interactively. Here we have implemented a switch to change between projection and slice view, a slider that adjusts the slice height $h$ and a numeric input to specify the center point $c$ explicitly. The display will update based on these inputs, in particular the view will jump to a new slice when $c$ is changed (though an interpolation might also be useful).

One might implement such an interactive interface via R Shiny, but in particular the tracking of the mouse position in small increments might pose a challenge. Here we found the dynamic graphics interface from Mathematica to be especially useful. In the following we first describe the relevant Mathematica functionalities and how they were used in the implementation, followed by a brief sketch of how the new approach might be implemented in the `tourr` package.

@@

Mathematica provides much utility and versatility, such as inbuilt data visualisation, data manipulation and analysis, dynamic functionality, and symbolic and numeric computation. Most of this inbuilt functionality is user-friendly and described in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there.

Importantly for our work, it is relatively simple  to create dynamic objects via the inbuilt commands `Manipulate` or `DynamicModule`. Control objects, such as sliders, locator panes, and input fields, can then be used on dynamic variables, and when there is a change in the dynamic variable the dynamic objects which contain that variable will be updated. This is the essential ingredient for this implementation of the manual tour.

However, it can be hard to alter the behaviour of the control objects such as setting the 1D slider on a different scale  (Slider increments add 10 points into the slice rather than increasing the slice height by 0.1 for example). \textcolor{red}{but this is no longer true if you replace it with animate??}
\textcolor{red}{Alex, is this still true with the controls you are using now?}

\textcolor{red}{Ursula: this comment about slider scale seems to specific, can we have a more general comment about difficulties, e.g. I remember some issues with the color mapping? Also there seem to be a lot of loops needed? Can you explain what the locator pane is?}

The examples presented in the penguins’ notebooks all use our primary new function, `SliceDynamic`. This function typically accepts grouped data in the form of a matrix where the second last column details the name of the group and the last column details the group index \textcolor{red}{(which can be one if there is only one group)}. After specifying the initial slice thickness and the slice range, the user is presented with an interactive display in which the control objects appear on the left and the slice visulisation appears on the right. The user can change the orientation of the slice via the locator pane, which changes the projection matrix;  a slider controls the slice thickness; and there is an input field, which changes the slice centre. The user can also change the appearance of the plot by zooming into the center or changing the point size with sliders provided.\textcolor{red}{this zoom doesn't work very well} The projected data can be displayed to contrast it with the slice by ticking a box. The explicit projection matrix can also be displayed via another checkbox. This is especially useful when we wish to import a projection that was identified as interesting in the manual exploration later, as shown in the applications below.

\textcolor{red}{Details about the implementation and usage instructions are given in the Appendix.}

### 2022-09-12 paper/paper.tex
Some work on the software section

XXX copied from the slicing part to be weaved in below:

The current implementation allows the user to individually enter the
components of the vector defining the center point, and updates the
display by jumping to the new slice.

The current implementation contains a switch to change between the slice
view and the projected view of the data, while keeping all other
settings fixed.

In practice we define the center point \(\mathbf{c}\) explicitly in
terms of all \(p\) components and calculate the slice display following
Eq. \ref{eq:slice}.

This was implemented as a slider, and to allow the user to efficiently
explore the interesting range we provide the slider range as an input.

@@

\begin{itemize}
\tightlist
\item
  \textbf{Why is this a good sandbox}
\item
  \textbf{Explain the functionality available in the notebooks}
\end{itemize}

Mathematica provides much utility and versatility, such as inbuilt data
visualisation, data manipulation and analysis, dynamic functionality,
and symbolic and numeric computation. Most of this inbuilt functionality
is quite user-friendly and can be found in the Mathematica
documentation; typically, numerous examples are provided within the
documentation, and some possible issues are outlined there. Also, it is
relatively simple to create dynamic objects via the inbuilt commands
\texttt{Manipulate} or \texttt{DynamicModule}. Control objects, such as
sliders, locator panes, and input fields, can then be used on dynamic
variables, and when there is a change in the dynamic variable the
dynamic objects which contain that dynamic variable will be updated.
This is the essential ingredient for this implementation of the manual
tour. However, it can be hard to alter the behaviour of the control
objects such as setting the 1D slider on a different scale (Slider
increments add 10 points into the slice rather than increasing the slice
height by 0.1 for example). \{\textcolor{red}Alex, is this still true
with the controls you are using now?\}

The examples presented in the penguins' notebooks all utilise our
primary new function, \texttt{SliceDynamic}. This function typically
accepts grouped data in the form of a matrix where the second last
column details the name of the group and the last column details the
group index. After specifying the initial slice height and the slice
range, the user is then presented with an interactive display in which
the control objects appear on the left and the slice visulisation
appears on the right. The user can change the orientation of the slice
via the locator pane, which changes the projection matrix; a slider
controls the slice height; and there is an input field, which changes
the slice centre. The user can also change the appearance of the plot by
zooming into the center or changing the point size with sliders
provided. The current projection can be displayed to contrast it with
the slice by ticking a box. The explicit projection matrix can be
displayed via another checkbox. The Slice Tour* notebooks include the
additional functions: ProjectionPlot, Projected2DSliderPlot, and
VisualiseSliceDyanmic. ProjectionPlot is very similar to SliceDynamic,
except it only displays projections and VisualiseSliceDynamic allows you
to discern which points exist inside the slice and outside it. However,
ProjectedLocatorPlot displays the interactive slightly differently.
Several locator panes are displayed above the plot and each one
corresponds to a row in the projection matrix but behaves the same as
ProjectionPlot. \{\textcolor{blue}(this may be easier to implement in
another language (?) )\}

>>>>

\hypertarget{structure}{%
\subsection{Structure}\label{structure}}

The implementation of the manual tour with slice control needs a
visualization of the current projection in terms of an axis display.
This display should be able to track the mouse position and adjust the
projection based on the user selection. The user can select one axis
(corresponding to one variable) via clicking on it, and then adjust the
position of that axis by dragging it during the click event. In practice
the closest axis will be selected, and the dragging results in a
tracking of the mouse position that will update the projection in small
steps, rendering interpolation unnecessary.

The interaction in the axis display need to be mapped back onto the
projection matrix, which in turn needs to be linked to both the
orthonormalization procedure and to a second display that is showing the
projected data.

In addition, we also want to be able to look at slices of the data and
select slicing parameters interactively. Here we have implemented a
switch to change between projection and slice view, a slider that
adjusts the slice height \(h\) and a numeric input to specify the center
point \(c\) explicitly. The display will update based on these inputs,
in particular the view will jump to a new slice when \(c\) is changed
(though an interpolation might also be useful).

One might implement such an interactive interface via R Shiny, but in
particular the tracking of the mouse position in small increments might
pose a challenge. Here we found the dynamic graphics interface from
Mathematica to be especially useful. In the following we first describe
the relevant Mathematica functionalities and how they were used in the
implementation, followed by a brief sketch of how the new approach might
be implemented in the \texttt{tourr} package.

@@

Mathematica provides much utility and versatility, such as inbuilt data
visualisation, data manipulation and analysis, dynamic functionality,
and symbolic and numeric computation. Most of this inbuilt functionality
is quite user-friendly and can be found in the Mathematica
documentation; typically, numerous examples are provided within the
documentation, and some possible issues are outlined there. Also, it is
relatively simple to create dynamic objects via the inbuilt commands
\texttt{Manipulate} or \texttt{DynamicModule}. Control objects, such as
sliders, locator panes, and input fields, can then be used on dynamic
variables, and when there is a change in the dynamic variable the
dynamic objects which contain that dynamic variable will be updated.
This is the essential ingredient for this implementation of the manual
tour. However, it can be hard to alter the behaviour of the control
objects such as setting the 1D slider on a different scale (Slider
increments add 10 points into the slice rather than increasing the slice
height by 0.1 for example).
\textcolor{red}{but this is no longer true if you replace it with animate??}
\textcolor{red}{Alex, is this still true with the controls you are using now?}

The examples presented in the penguins' notebooks all utilise our
primary new function, \texttt{SliceDynamic}. This function typically
accepts grouped data in the form of a matrix where the second last
column details the name of the group and the last column details the
group index
\textcolor{red}{(which can be one if there is only one group)}. After
specifying the initial slice thickness and the slice range, the user is
then presented with an interactive display in which the control objects
appear on the left and the slice visulisation appears on the right. The
user can change the orientation of the slice via the locator pane, which
changes the projection matrix; a slider controls the slice thickness;
and there is an input field, which changes the slice centre. The user
can also change the appearance of the plot by zooming into the center or
changing the point size with sliders
provided.\textcolor{red}{this zoom doesn't work very well} The current
projection can be displayed to contrast it with the slice by ticking a
box. The explicit projection matrix can also be displayed via another
checkbox in order to store it for further study.
